### 2023-10-27 abstract.tex
Some final tweaks and changelist

The marked lambda calculus is bidirectionally typed, so localization decisions are systematically predictable based on a local flow of typing information. 
Constraint-based type inference can bring more distant information to bear in discovering inconsistencies but this notoriously complicates error localization. We approach this problem by deploying constraint solving as a type-hole-filling layer atop this gradual bidirectionally typed core. Errors arising from inconsistent unification constraints are localized exclusively to type and expression holes, i.e. the system identifies unfillable holes using a sytem of traced  provenances, rather than localized in an \emph{ad hoc} manner to particular expressions. The user can then interactively shift these errors to particular downstream expressions by selecting from suggested partially consistent type hole fillings, which returns control back to the bidirectional system. We implement this type hole inference system in Hazel. 
\end{abstract}

>>>>

The marked lambda calculus is bidirectionally typed, so localization decisions are systematically predictable based on a local flow of typing information. 
Constraint-based type inference can bring more distant information to bear in discovering
inconsistencies but this notoriously complicates error localization. We approach this problem by
deploying constraint solving as a type-hole-filling layer atop this gradual bidirectionally typed
core. Errors arising from inconsistent unification constraints are localized exclusively to type and
expression holes, i.e., the system identifies unfillable holes using a system of traced  provenances, rather than localized in an \emph{ad hoc} manner to particular expressions. The user can then interactively shift these errors to particular downstream expressions by selecting from suggested partially consistent type hole fillings, which returns control back to the bidirectional system. We implement this type hole inference system in Hazel. 
\end{abstract}

### 2023-10-27 calculus.tex
Some final tweaks and changelist

But what happens when the synthesized type of an expression is \emph{not} consistent with the
expected type, i.e. that the premise $\consistent{\TMV}{\TMV'}$ fails? Recalling the example of
\cref{fig:calculus-examples-inconsistent-types}, subsumption would be used to analyze $\ETrue$
against $\TNum$ when checking $\EPlus{\ETrue}{1}$, which fails since $\inconsistent{\TNum}{\TBool}$, i.e.
\textsc{UASubsume} does not apply. At this point, traditional typing semantics would simply fail the type
checking process. \textsc{MKASubsume} would also not be applicable, leaving marking
undefined in those cases.

@@

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\
    \ctxAnaTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxAnaTypeU{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
  }{UALam}
\end{mathpar}
%
The synthesis rule is standard. Analysis employs the judgment $\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$,
which establishes that $\TMV$ is a \emph{matched arrow type} \cite{cimini2016}, i.e. it may be
considered an arrow type. Defined in \cref{fig:calculus-matched-arrow}, this notion is purely a
technical mechanism to avoid duplication of rules related to arrow types \cite{siek2015}.

@@

Another error arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
$\inconsistent{\TMV}{\TMV_1}$, i.e. the actual type annotation for $x$ is inconsistent with the
expected input type. The \emph{inconsistent ascription mark} $\ECLamInconAsc{x}{\TMV}{\ECMV}$ indicates exactly this
error, and we add a final pair of analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \inconsistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLamInconAsc{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam3}

@@

Earlier versions of Hazel achieved total type error recovery by using a term-based structure editor, which maintained delimiter matching by construction, albeit at the cost of some syntactic rigidity.
To specify such a structure editor formally, \citet{omar2017b} introduced the Hazelnut action calculus, which operates over a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{cursor}.
The synthetic action judgment, written $\ASEAction{\ctx}{\ZMV}{\TMV}{\ZMV'}{\TMV'}{\AMV}$, performs some edit action $\AMV$ on some \emph{zippered expression} $\ZMV$ (which is an expression with a superimposed cursor, denoted $\hat{\EMV}$ in the original work) that synthesizes type $\TMV$, producing some new $\ZMV'$ that synthesizes types $\TMV'$.
Type and binding information is propagated to the location of the cursor during these operations, allowing them to insert the necessary error marks (i.e. non-empty holes, in Hazelnut's parlance).
When, for example, wrapping an expression whose type is inconsistent with $\TNum$ in an arithmetic operation, the calculus inserts an error mark around that operand:
%
\begin{mathpar}
    \inferrule{
        \inconsistent{\TMV}{\TNum}
    }{
        \ASCon{\ctx}{\ZCursor{\EMV}}{\TMV}{\ZPlusL{\ECInconType{\ECMV}}{\ZCursor{\EEHole}}}{\TNum}{\SPlusL}
    }
\end{mathpar}

>>>>

But what happens when the synthesized type of an expression is \emph{not} consistent with the
expected type, i.e., that the premise $\consistent{\TMV}{\TMV'}$ fails? Recalling the example of
\cref{fig:calculus-examples-inconsistent-types}, subsumption would be used to analyze $\ETrue$
against $\TNum$ when checking $\EPlus{\ETrue}{1}$, which fails since
$\inconsistent{\TNum}{\TBool}$---\textsc{UASubsume} does not apply. At this point, traditional typing semantics would simply fail the type
checking process. \textsc{MKASubsume} would also not be applicable, leaving marking
undefined in those cases.

@@

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\
    \ctxAnaTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxAnaTypeU{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
  }{UALam}
\end{mathpar}
%
The synthesis rule is standard. Analysis employs the judgment $\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$,
which establishes that $\TMV$ is a \emph{matched arrow type} \cite{cimini2016}, i.e., it may be
considered an arrow type. Defined in \cref{fig:calculus-matched-arrow}, this notion is purely a
technical mechanism to avoid duplication of rules related to arrow types \cite{siek2015}.

@@

Another error arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
$\inconsistent{\TMV}{\TMV_1}$, i.e., the actual type annotation for $x$ is inconsistent with the
expected input type. The \emph{inconsistent ascription mark} $\ECLamInconAsc{x}{\TMV}{\ECMV}$ indicates exactly this
error, and we add a final pair of analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \inconsistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLamInconAsc{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam3}

@@

Earlier versions of Hazel achieved total type error recovery by using a term-based structure editor, which maintained delimiter matching by construction, albeit at the cost of some syntactic rigidity.
To specify such a structure editor formally, \citet{omar2017b} introduced the Hazelnut action calculus, which operates over a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{cursor}.
The synthetic action judgment, written $\ASEAction{\ctx}{\ZMV}{\TMV}{\ZMV'}{\TMV'}{\AMV}$, performs an edit action $\AMV$ on some \emph{zippered expression} $\ZMV$ (which is an expression with a superimposed cursor, denoted $\hat{\EMV}$ in the original work) that synthesizes type $\TMV$, producing some new $\ZMV'$ that synthesizes types $\TMV'$.
Type and binding information is propagated to the location of the cursor during these operations, allowing them to insert the necessary error marks (i.e. non-empty holes, in Hazelnut's parlance).
When, for example, wrapping an expression whose type is inconsistent with $\TNum$ in an arithmetic operation, the calculus inserts an error mark around that operand:
%
\begin{mathpar}
    \inferrule{
        \inconsistent{\TMV}{\TNum}
    }{
        \ASCon{\ctx}{\ZCursor{\EMV}}{\TMV}{\ZPlusL{\ECInconType{\ECMV}}{\ZCursor{\EEHole}}}{\TNum}{\SPlusL}
    }
\end{mathpar}

### 2023-10-27 intro.tex
Some final tweaks and changelist

These service gaps are fundamentally rooted in a definitional gap: a type system defined in the conventional way, 
e.g. in the tradition of the typed lambda calculus and its derivatives \cite{pierce2002},
assigns meaning only to well-typed programs. 
If a type error appears \emph{anywhere}, the program is formally meaningless \emph{everywhere}.

This gap problem has prompted considerable practical interest in 
(1)~\textbf{type error localization}: mechanisms for identifying the location(s) in a program that explain a type error, and 
(2) \textbf{type error recovery}: mechanisms that allow the system to optimistically recover from a localized type error 
and continue on to locate other errors and provide downstream semantic services, 
ideally at every location in the program and with minimal degradation in service.
Essentially all widely-used programming systems have some support for type error localization, 
e.g. in compiler error messages 
or directly in the editor via markings decorating localized errors. Developers are known to attend to reported error locations when debugging type errors \cite{joosten1993}.
Many systems also attempt recovery in certain situations, discussed below. 
However, type error localization and recovery mechanisms have developed idiosyncratically, 
in part as folklore amongst language and tool implementors. 
Different type checkers or language servers \cite{barros2022,bour2018}, even for the same language, localize and recover from type errors in different ways, 
with little in the way of unifying theory of the sort that grounds the design of modern type systems themselves.

@@

To recover further, we might choose to also ignore the bidirectional type system's demand that the conditional guard be confirmed to be a boolean expression 
(because the type of \li{f(y)} is also not well-defined) 
and continue into its branches, observing that they have inconsistent types, \li{Bool} and \li{Int}. 
There are several ways to localize this error. 
One approach, taken e.g. by Elm and \li{merlin}, would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
Heuristics have been developed to help make this choice less arbitrary, e.g. 
by observing recent editing history \cite{chuchem2019}, 
or training a machine learning model \cite{seidel2017}. 
A less \emph{ad hoc} approach, which Hazel takes above, is to localize the inconsistency to the conditional as a whole, reporting that the problem is that the branch types differ. When the cursor is on this conditional expression, the Hazel type inspector---a status bar that reports typing information, including error messages, about the term at the cursor \cite{potter2020}---displays:

@@

\Cref{sec:calculus} of this paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus}.
We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
type annotations to expressions \cite{pierce2000,dunfield2019}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{siek2006,siek2015}, and 
(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{omar2017b}, operationalizing the ``red lines'' that an editor displays.
The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solarlezama2013}) can be marked, i.e. its errors can be simultaneously localized, such that the resulting program has a well-defined type and binding structure.
We establish this and other properties as metatheorems that we mechanize in the Agda proof assistant. 
As we define the calculus, we consider a number of situations where error localization decisions are subtle and conclude by defining some extensions of the intentionally minimal core calculus, e.g. with destructuring patterns and System F-style polymorphism, 
that serve as case studies of the general approach that we hope will be adopted by language designers defining practical type systems.

@@

For example, in the following Hazel expression (ignoring the error on the type hole for the moment), bidirectional error localization 
does not mark any uses of \li{x} as erroneous because a type for \li{x} cannot locally be inferred (the let-bound expression is an empty hole, which has unknown type). However, by gathering constraints from the three uses of \li{x} and attempting to unify, we see that the type hole is unfillable. Rather than arbitrarily privileging one of these uses, as is the case in languages like OCaml (and which has led to the development of complex heuristics, e.g. using machine learning \cite{seidel2017}, to avoid misleading programmers), this error is localized, with a \li{!}, to the type hole itself, with the error message providing partially consistent solutions as shown.
\\
\begin{center}
    % \includegraphics[scale=0.55]{images/intro-thi.png}
    % \includegraphics[scale=0.4]{images/example_CI.png}
    \includegraphics[scale=0.4]{images/figSugg.png}
\end{center}

The user can interactively explore possible error localizations by hovering over a partially consistent suggestion, which (temporarily) fills the type hole and thus returns the localization decision to the bidirectional type system.
Here, each choice causes a different set of two uses of \li{x} to be marked, e.g. the first and third use for \li{Int -> Int} (shown above). Until the choice is finalized, the type hole continues to operate gradually.
\cref{sec:thi} describes this distinctively neutral approach to blending local and constraint-based
type inference in Hazel, where bidirectional typing is used to systematically mark erroneous expressions in the program and constraint-based inference (using entirely standard algorithms, which we do not repeat in this paper) is used exclusively to locate unfillable holes.

>>>>

These service gaps are fundamentally rooted in a definitional gap: a type system defined in the conventional way, 
e.g., in the tradition of the typed lambda calculus and its derivatives \cite{pierce2002},
assigns meaning only to well-typed programs. 
If a type error appears \emph{anywhere}, the program is formally meaningless \emph{everywhere}.

This gap problem has prompted considerable practical interest in 
(1)~\textbf{type error localization}: mechanisms for identifying the location(s) in a program that explain a type error, and 
(2) \textbf{type error recovery}: mechanisms that allow the system to optimistically recover from a localized type error 
and continue on to locate other errors and provide downstream semantic services, 
ideally at every location in the program and with minimal degradation in service.
Essentially all widely-used programming systems have some support for type error localization, 
e.g., in compiler error messages 
or directly in the editor via markings decorating localized errors. Developers are known to attend to reported error locations when debugging type errors \cite{joosten1993}.
Many systems also attempt recovery in certain situations, discussed below. 
However, type error localization and recovery mechanisms have developed idiosyncratically, 
in part as folklore amongst language and tool implementors. 
Different type checkers or language servers \cite{barros2022,bour2018}, even for the same language, localize and recover from type errors in different ways, 
with little in the way of unifying theory of the sort that grounds the design of modern type systems themselves.

@@

To recover further, we might choose to also ignore the bidirectional type system's demand that the conditional guard be confirmed to be a boolean expression 
(because the type of \li{f(y)} is also not well-defined) 
and continue into its branches, observing that they have inconsistent types, \li{Bool} and \li{Int}. 
There are several ways to localize this error. 
One approach, taken e.g., by Elm and \li{merlin}, would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
Heuristics have been developed to help make this choice less arbitrary, e.g., 
by observing recent editing history \cite{chuchem2019}, 
or training a machine learning model \cite{seidel2017}. 
A less \emph{ad hoc} approach, which Hazel takes above, is to localize the inconsistency to the conditional as a whole, reporting that the problem is that the branch types differ. When the cursor is on this conditional expression, the Hazel type inspector---a status bar that reports typing information, including error messages, about the term at the cursor \cite{potter2020}---displays:

@@

\Cref{sec:calculus} of this paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus}.
We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
type annotations to expressions \cite{pierce2000,dunfield2019}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{siek2006,siek2015}, and 
(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{omar2017b}, operationalizing the ``red lines'' that an editor displays.
The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solarlezama2013}) can be marked, i.e., its errors can be simultaneously localized, such that the resulting program has a well-defined type and binding structure.
We establish this and other properties as metatheorems that we mechanize in the Agda proof assistant. 
As we define the calculus, we consider a number of situations where error localization decisions are
subtle and conclude by defining some extensions of the intentionally minimal core calculus, e.g., with destructuring patterns and System F-style polymorphism, 
that serve as case studies of the general approach that we hope will be adopted by language designers defining practical type systems.

@@

For example, in the following Hazel expression (ignoring the error on the type hole for the moment), bidirectional error localization 
does not mark any uses of \li{x} as erroneous because a type for \li{x} cannot locally be inferred
(the let-bound expression is an empty hole, which has unknown type). However, by gathering
constraints from the three uses of \li{x} and attempting to unify, we see that the type hole is
unfillable. Rather than arbitrarily privileging one of these uses, as is the case in languages like
OCaml (and which has led to the development of complex heuristics, e.g., using machine learning \cite{seidel2017}, to avoid misleading programmers), this error is localized, with a \li{!}, to the type hole itself, with the error message providing partially consistent solutions as shown.
\\
\begin{center}
    % \includegraphics[scale=0.55]{images/intro-thi.png}
    % \includegraphics[scale=0.4]{images/example_CI.png}
    \includegraphics[scale=0.4]{images/figSugg.png}
\end{center}

The user can interactively explore possible error localizations by hovering over a partially consistent suggestion, which (temporarily) fills the type hole and thus returns the localization decision to the bidirectional type system.
Here, each choice causes a different set of two uses of \li{x} to be marked, e.g., the first and third use for \li{Int -> Int} (shown above). Until the choice is finalized, the type hole continues to operate gradually.
\cref{sec:thi} describes this distinctively neutral approach to blending local and constraint-based
type inference in Hazel, where bidirectional typing is used to systematically mark erroneous expressions in the program and constraint-based inference (using entirely standard algorithms, which we do not repeat in this paper) is used exclusively to locate unfillable holes.

### 2023-10-27 polymorphism.tex
Some final tweaks and changelist

We leave as future work the task of defining marked versions of even more elaborate bidirectional
type systems, e.g. \citet{dunfield2013}'s rather substantial formulation of implicit type
application in a bidirectional setting (which would have to handle, for example, the situation where
no implicit argument can be resolved) or \citet{lennonbertrand2022}'s gradual bidirectional
variant of the dependently typed calculus of constructions.
Languages with operator overloading may provide interesting localization decisions.
So that overloading can be resolved during typing and marking, the modes of one or both operand must be changed to synthesis.
If the operator is not defined given the synthesized operand type(s), an error might be marked on the operator itself.
Type classes \cite{wadler1989} present a symmetric alternative where the operands may stay in analytic mode.
In either approach, the necessary error marks and localization decisions may be arrived at via the recipe we have described.

>>>>

We leave as future work the task of defining marked versions of even more elaborate bidirectional
type systems, e.g., \citet{dunfield2013}'s rather substantial formulation of implicit type
application in a bidirectional setting (which would have to handle, for example, the situation where
no implicit argument can be resolved) or \citet{lennonbertrand2022}'s gradual bidirectional
variant of the dependently typed calculus of constructions.
Languages with operator overloading may provide interesting localization decisions.
So that overloading can be resolved during typing and marking, the modes of one or both operand must be changed to synthesis.
If the operator is not defined given the synthesized operand type(s), an error might be marked on the operator itself.
Type classes \cite{wadler1989} present a symmetric alternative where the operands may stay in analytic mode.
In either approach, the necessary error marks and localization decisions may be arrived at via the recipe we have described.

### 2023-10-27 related.tex
Some final tweaks and changelist

In addition, there has been considerable work on improving explanations for type errors, e.g. by providing sample inputs (dynamic witnesses) that elicit runtime errors \cite{seidel2016}. With this approach, one can generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with around 70\% accuracy.
These and a variety of other approaches to explaining errors are a fruitful avenue for future integration into the system.

>>>>

In addition, there has been considerable work on improving explanations for type errors, e.g., by providing sample inputs (dynamic witnesses) that elicit runtime errors \cite{seidel2016}. With this approach, one can generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with around 70\% accuracy.
These and a variety of other approaches to explaining errors are a fruitful avenue for future integration into the system.
